Double Robust, Flexible Adjustment Methods for Causal Inference: An Overview and an Evaluation

### 2024-03-11 draft/ms-adj-draft.Rmd
full draft for working groups

---
output:
  # bookdown::word_document2:
    # reference_docx: "word-template.docx"
  bookdown::pdf_document2:
    toc: no
    number_sections: yes
    pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
    latex_engine: xelatex
always_allow_html: true
header-includes:
  #- \usepackage{setspace}\doublespace
  # - \usepackage[nolists, fighead, tabhead]{endfloat}
  # - \usepackage{endnotes}
  # - \let\footnote=\endnote
  # - \setlength{\headheight}{14.5pt}
  # - \setlength{\headheight}{13.6pt}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \lhead{N.I. Hoffmann}
  - \rhead{`r format(Sys.time(), '%B %e, %Y')`}
# - \newcommand{\beginsupplement}{\setcounter{table}{0}  
# \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} 
# \renewcommand{\thefigure}{A\arabic{figure}}}

@@

citeproc: no
fontfamily: mathpazo
#fontsize: 11pt
# geometry: margin=.6in
indent: yes
link-citations: yes
linkcolor: blue
lang: 'en-US'

@@

title: "Double Robust, Flexible Adjustment Methods for Causal Inference: An Overview and an Evaluation"
# subtitle: "American Sociological Association Annual Meeting 2024"
author:  Nathan I. Hoffmann, Department of Sociology, UCLA
date: "`r format(Sys.time(), '%B %e, %Y')`"

abstract: "Double robust methods for flexible covariate adjustment in causal inference have proliferated in recent years. Despite their apparent advantages, these methods remain underutilized by social scientists. It is also unclear whether these methods actually outperform more traditional methods in finite samples. This paper has two aims: It is a guide to some of the latest methods in double robust, flexible covariate adjustment for causal inference, and it compares these methods to more traditional statistical methods. It does this by using both simulated data where the treatment effect estimate is known, and then using comparisons of experimental and observational data from the National Supported Work Demonstration. Methods covered include Augmented Inverse Probability Weighting, Targeted Maximum Likelihood Estimation, and Double/Debiased Machine Learning. Results suggest that these methods do not necessarily outperform OLS regression or matching on propensity score estimated by logistic regression, even in cases where the data generating process is not linear."

@@

library(SuperLearner)
library(broom)
library(knitr)
library(kableExtra)
library(here)
library(haven)
library(tidyverse)

@@

## add lalonde ####
# read_csv(here('files_combined', 'lm_df.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde', 'lm_df.csv'))) %>%
#   write_csv(here('files', 'lm_df.csv'))
# 
# 
# read_csv(here('files_combined', 'psm_df.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde', 'psm_df.csv'))) %>%
#   write_csv(here('files', 'psm_df.csv'))
# 
# read_csv(here('files_combined', 'lin.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde', 'lin.csv'))) %>%
#   write_csv(here('files', 'lin.csv'))
# 
# read_csv(here('files_combined', 'gcomp.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde', 'gcomp.csv'))) %>%
#   write_csv(here('files', 'gcomp.csv'))
# 
# read_csv(here('files_combined/homemade', 'aipw.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde/homemade', 'aipw.csv'))) %>%
#   write_csv(here('files/homemade', 'aipw.csv'))
# 
# read_csv(here('files_combined/homemade', 'aipw_trunc.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde/homemade', 'aipw_trunc.csv'))) %>%
#   write_csv(here('files/homemade', 'aipw_trunc.csv'))
# 
# read_csv(here('files_combined/homemade', 'tmle.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde/homemade', 'tmle.csv'))) %>%
#   write_csv(here('files/homemade', 'tmle.csv'))
# 
# read_csv(here('files_combined/homemade', 'dml_ols_logit.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde/homemade', 'dml_ols_logit.csv'))) %>%
#   write_csv(here('files/homemade', 'dml_ols_logit.csv'))
# 
# read_csv(here('files_combined/homemade', 'dml_grf.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde/homemade', 'dml_grf.csv'))) %>%
#   write_csv(here('files/homemade', 'dml_grf.csv'))
# 
# read_csv(here('files_combined/homemade', 'dml_superlearner.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde/homemade', 'dml_superlearner.csv'))) %>% 
#   write_csv(here('files/homemade', 'dml_superlearner.csv'))
# 
# 
# readRDS(here('files_combined/homemade', 'ols_logit_pred.RDS')) %>%
#   bind_rows(readRDS(here('files_lalonde/homemade', 'ols_logit_pred.RDS'))) %>%
#   saveRDS(here('files/homemade', 'ols_logit_pred.RDS'))
# 
# readRDS(here('files_combined/homemade', 'grf_pred.RDS')) %>%
#   bind_rows(readRDS(here('files_lalonde/homemade', 'grf_pred.RDS'))) %>%
#   saveRDS(here('files/homemade', 'grf_pred.RDS'))
# 
# readRDS(here('files_combined/homemade', 'superlearner_pred.RDS')) %>%
#   bind_rows(readRDS(here('files_lalonde/homemade', 'superlearner_pred.RDS'))) %>%
#   saveRDS(here('files/homemade', 'superlearner_pred.RDS'))

```{r sim-homemade, eval = F}
sim_homemade <- function(n, m = NULL){
  if(is.null(m)){
    m <- round(n/100)
  }

  dat <- replicate(m, rnorm(n, 0, 5)) %>% 
    as.data.frame() 
  dat_original <- dat

  # first make treatment indicator
  dat <- data.frame(1, dat)
  coefs <- rnorm(m+1, 0, 2)
  d <- round(1/(1+exp(-as.matrix(dat) %*% as.matrix(coefs))))

  dat <- data.frame(dat, d)

  # Set treatment effect to 2
  coefs <- c(rnorm(m+1, 0, 2), 2)
  y <- as.matrix(dat) %*% as.matrix(coefs) + rnorm(1)

  dat <- data.frame(dat, y)

  dat_new <- dat_original

  transformers <- rnorm(m, 0, 2)

  for(i in 1:m){
    dat_new[,i] <- dat_new[,i] * transformers[i]
  }

  # m/4 squared terms
  dat_new <- dat_new %>% 
    mutate(across(sample(1:m, m/4, replace = F), function(x) x^2))

  # m/4 exponentiated 
  dat_new <- dat_new %>% 
    mutate(across(sample(1:m, m/4, replace = F), function(x) exp(x)))

  # m/4 sin 
  dat_new <- dat_new %>% 
    mutate(across(sample(1:m, m/4, replace = F), function(x) sin(x)))

  # m/2 2-way interactions
  for(i in 1:round(m/2)){
    index <- sample(1:m, 2, replace = F)
    dat_new[, index[1]] <- dat_new[,index[1]] * dat_new[,index[2]]
  }

  # m/4 3-way interactions
  for(i in 1:round(m/4)){
    index <- sample(1:m, 3, replace = F)
    dat_new[, index[1]] <- dat_new[,index[1]] * dat_new[,index[2]] * dat_new[,index[3]]
  }

  dat_final <- data.frame(dat_new, d = dat[,'d'], y = dat[,'y'])
  return(dat_final)
}

# 500, 1000, 5000, 10k, 50k, 100k
# Set m = 1% of n
# 50 of each 
# then set n = 5000 and m = 10, 50, 200, 1000
set.seed(185)
sim_homemade_list <- list()
for(n in c(500, 1000, 5000, 1e3, 5e3, 1e5)){
  print(paste0('n = ', n))
  for(i in 1:50){
    print(i)
    sim_homemade_list[[paste0(n, i)]] <- sim_homemade(n) %>%
      mutate(n = n, m  = n/100, dataset = i, simulation = 'n')
  }
}
for(m in c(10, 50, 200, 500, 1000)){
  print(paste0('m = ', m))
  for(i in 1:50){
    print(i)
    sim_homemade_list[[paste0(n, i)]] <- sim_homemade(n = n, m = m) %>%
      mutate(n = n, m  = m, dataset = i, simulation = 'm')
  }
}

saveRDS(sim_homemade_list, here('files', 'sim_homemade.RDS'))

```{r github-datageneration, eval = F}
# from https://github.com/QuantLet/DataGenerationForCausalInference

### This generates the Simulation Data as a dataframe
# by Daniel Jacob (daniel.jacob@hu-berlin.de)

# Arguments to specify are: 
# Y = Outcome (dependend variable). Either continuous or binary. 
# N = Number of observations (real number)
# k = Number of covariates (real number). At least 10 
# random_d = treatment assignment: (Either T for random assignment or F for confounding on X)
# theta = treatment effect: (Either real number for only one theta, or "binary" {0.1,0.3}, "con" for continuous values (0.1,0.3) or "big" for {1,0.4})
# var = Size of the variance (Noise-level)

#Required Packages
if(!require("clusterGeneration")) install.packages("clusterGeneration"); library("clusterGeneration")
if(!require("mvtnorm")) install.packages("mvtnorm"); library("mvtnorm")

datagen <- function(N,y,k,random_d,theta,var) {

  N = N
  k = k
  b = 1 / (1:k)
  # = Generate covariance matrix of z = #
  sigma <- genPositiveDefMat(k, "unifcorrmat")$Sigma
  sigma <- cov2cor(sigma)

  z <- rmvnorm(N, sigma = sigma) # = Generate z = #

  ### Options for D (m_(X))
  if (random_d == T) {
    d <- rep(c(0, 1), length.out = N)
  } else {
    d_prop <- pnorm(z %*% b) # D is dependent on Za
    d <- as.numeric(rbinom(N, prob = d_prop, size = 1))
  }

  ### Options for theta
  if (theta == "con") {
    theta_s <- as.vector(sin(z %*% b) ^ 2)
    theta <-
      (theta_s - min(theta_s)) * (0.3 - 0.1) / (max(theta_s) - min(theta_s)) +
      0.1
  } else if (theta == "binary") {
    theta_low <-
      rbinom(N, pnorm((z[, 6] * (z[, 1] %*% t(
        z[, 5]
      )) * z[, 2]) ^ 2), size = 1)
    theta <-
      ifelse(theta_low == 1, 0.3, 0.1)
  } else if (theta == "big") {
    theta_big <-
      rbinom(N, pnorm((z[, 6] * (z[, 1] %*% t(
        z[, 5]
      )) * z[, 2]) ^ 2), size = 1)
    theta <- ifelse(theta_big == 1, 1, 0.4)
  }  else {
    theta == theta
  }

  g <- as.vector(cos(z %*% b) ^ 2)

  if(y=="binary") {
    y1 <- theta * d + g 
    y1.1 <- rbinom(N,prob=pnorm(scale(y1)),size=1)
    y <- y1.1
  } else {y <- theta * d + g + rnorm(N,0,var)}

  data <- as.data.frame(y)
  data <- cbind(data, theta, d, z)
  colnames(data) <- c("y", "theta", "d", c(paste0("V", 1:k)))

  return(data)
}

### Example
dataset <- datagen(y="binary",N = 2000, k = 20, random_d = F, theta = "binary", var = 1)
summary(dataset)
str(dataset)

glm(y ~ ., data = dataset, family = binomial(link = 'logit')) %>%
  summary()
```

```{r trash, eval = F}
dat_transform <- function(dat){
  # first make treatment indicator
  coefs <- runif(ncol(dat), -1, 1)

  # m/4 squared terms
  dat <- dat %>% 
    mutate(across(sample(1:m, m/4, replace = F), function(x) x^2))

  # m/4 exponentiated 
  dat <- dat %>% 
    mutate(across(sample(1:ncol(dat), m/4, replace = F), function(x) exp(x)))

  # m/4 sin 
  dat <- dat %>% 
    mutate(across(sample(1:ncol(dat), m/4, replace = F), function(x) sin(x)))

  # m/2 2-way interactions
  dat <- bind_cols(dat,
    replicate(m/2, {
      index <- sample(1:m, 2, replace = F)
      dat[,index[1]] * dat[,index[2]]
      }, simplify = T))

  # m/4 3-way interactions
  dat <- bind_cols(dat,
    replicate(m/4, {
      index <- sample(1:m, 3, replace = F)
      dat[,index[1]] * dat[,index[2]] * dat[,index[3]]
      }, simplify = T))

  coefs <- runif(ncol(dat), -1, 1)

  y <- as.matrix(dat) %*% as.matrix(coefs)

  return(y)
}

sim_homemade <- function(m, n){
  m <- m-2
  dat <- replicate(m, rnorm(n, 0, 5)) %>%
    cbind(replicate(m/4, rbinom(n, 1, .5))) %>%
    as.data.frame()
  dat_original <- dat

  # first make treatment indicator
  dat <- data.frame(1, dat)
  coefs <- rnorm(ncol(dat), 0, 2)
  d <- round(1/(1+exp(-as.matrix(dat) %*% as.matrix(coefs))))

  dat <- data.frame(dat, d)

  # Set treatment effect to 2
  coefs <- c(rnorm(m+1, 0, 2), 2)
  y <- as.matrix(dat) %*% as.matrix(coefs) + rnorm(1)

  dat <- data.frame(dat, y)

  dat_new <- dat_original

  transformers <- rnorm(m, 0, 2)

  for(i in 1:m){
    dat_new[,i] <- dat_new[,i] * transformers[i]
  }

  # m/4 squared terms
  dat_new <- dat_new %>% 
    mutate(across(sample(1:m, m/4, replace = F), function(x) x^2))

  # m/4 exponentiated 
  dat_new <- dat_new %>% 
    mutate(across(sample(1:ncol(m), m/4, replace = F), function(x) exp(x)))

  # m/4 sin 
  dat_new <- dat_new %>% 
    mutate(across(sample(1:ncol(m), m/4, replace = F), function(x) sin(x)))

  # m/2 2-way interactions
  for(i in 1:round(m/2)){
    index <- sample(1:m, 2, replace = F)
    dat_new[, index[1]] <- dat_new[,index[1]] * dat_new[,index[2]]
  }

  # m/4 3-way interactions
  for(i in 1:round(m/4)){
    index <- sample(1:m, 3, replace = F)
    dat_new[, index[1]] <- dat_new[,index[1]] * dat_new[,index[2]] * dat_new[,index[3]]
  }

  dat_final <- data.frame(dat_new, d = dat[,'d'], y = dat[,'y'])
  return(dat_final)
}

@@

@@

<!-- # Abstract -->

<!-- This thesis aims to compare different advanced adjustment methods. It will compare Imben & Rubins propensity matching, stabilized IPWs, TMLE, double machine learning (Chernozhukov et al 2018), Ratkovic (2021), and BART. -->

@@

Statistical methods for flexible covariate adjustment in causal inference have proliferated in recent years. These methods have a number of strengths over traditional regression methods: They make few functional form assumptions, can accommodate large numbers of covariates, and produce easily interpretable treatment effect estimates. Many of these methods also have a "double robust" property: They estimate one model for the treatment exposure and another for the outcome, and as long as at least one is correctly specified, then the treatment effect will be estimated consistently. Despite their apparent advantages, these methods remain underutilized by social scientists. Part of the barrier has been lack of familiarity with these methods. It has also been unclear how these methods compare, or whether such methods actually perform better than traditional methods in finite samples. 
<!-- And when researchers want to use an existing R package for these methods, the array of possible estimation methods and other options can be dizzying.   -->

This paper makes advances on these fronts. First, it is a guide to some of the latest methods in double robust, flexible covariate adjustment for causal inference, explaining the methods to a social scientist audience. Second, it compares these methods to more traditional statistical methods using both simulations [@dorie_2019_automated] and the National Support for Work Demonstration (NSW) originally analyzed by @lalonde_1986_evaluating.

Methods covered include Augmented Inverse Probability Weighting [AIPW, @glynn_2010_introduction], Targeted Maximum Likelihood Estimation [TMLE, @vanderlaan_2006_targeted], and Double or Debiased Machine Learning [DML, @chernozhukov_2018_double]. This paper reviews the theory behind these methods as well as simple R implementations of them on simulations and real data. These methods are compared to two methods commonly used by social scientists: ordinary least squares (OLS) regression, and matching on propensity scores estimated from logistic regression (PSM).

@@

Although previous work has compared different methods for covariate adjustment, there has not yet been a thorough evaluation and comparison of the recent and popular double robust methods of AIPW, TMLE, and DML. @austin_2014_comparison compares methods for matching on the propensity score. @dorie_2019_automated report the results for the Atlantic Causal Inference Conference 2016 competition. Some of the methods used incorporate TMLE but not AIPW or DML.  @cousineau_2022_estimating evaluates the performance of optimization-baseds methods for causal inference, but these do not include the double robust methods covered in the current paper. @kang_2007_demystifying provide an overview and evaluation of double robust methods in the context of missing data. The authors consider AIPW but not TMLE or DML. @chatton_2020_gcomputation compare four methods: G-computation, IPW, full matching, and TMLE. These authors only consider one double robust method, and their focus is on omitted variable bias rather than determining which method is the most useful. @knaus_2022_double reviews double robust methods in an econometrics setting but does not compare them to traditional statistical methods for covaraite adjustment.

@@

According to @bang_2005_doubly, double robust methods have their origins in missing data models. @robins_1994_estimation and @rotnitzky_1998_semiparametric developed augmented orthogonal inverse probability-weighted (AIPW) estimators in missing data models, and @scharfstein_1999_adjusting showed that AIPW was double robust and extended to causal inference (since causal inference is fundamentally a missing data problem).

But @kang_2007_demystifying argue that double robust methods are older. They cite work by @cassel_1976_results, who proposed “generalized regression estimators” for population means from surveys where sampling weights must be estimated. Arguably, double robust methods go back even further than this. The form of double robust methods is similar to residual-on-residual regression, which dates back to @frisch_1933_partial famous FWL theorem:

$$\beta_D = \frac{\text{Cov}(\tilde Y_i, \tilde D_i)}{\text{Var}(\tilde D_i)}$$

where $\tilde D_i$ is the residual part of $D_i$ after regressing it on $X_i$, and $\tilde Y_i$ is the residual part of $Y_i$ after regressing it on $X_i$. This formulation writes the regression coefficient as composed of an outcome model ($\tilde Y_i$) and exposure model ($\tilde D_i$), the two models used in double robust estimators.

There are also links between double robust methods and matching with regression adjustment. This work goes back to at least @rubin_1973_use, who suggested that regression adjustment in matched data produces less biased estimates that either matching (exposure adjustment) or regression (outcome adjustment) do by themselves.

Today, double robust methods abound [e.g. @ratkovic_2023_relaxing; @sloczynski_2018_general; @arkhangelsky_2021_doublerobust; @dukes_2022_doubly; @kennedy_2023_semiparametric]. This paper focuses on three of the more popular or fundamental methods.

## Aims of Double Robust Methods
Double robust methods for covariate adjustment aim to overcome what many consider to be the downsides of both traditional statistical methods and flexible machine learning methods. Statistical methods that are popular with social scientists -- such as OLS regression and matching on propensity scores from logistic regression -- have two main weaknesses that double robust methods address. First, they assume simple (linear or transformed linear) functional forms. In the presence of highly nonlinear data generating processes, they may provide biased estimates. Second, these methods cannot handle large numbers of covariates relative to sample size. While some machine learning methods can produce estimates even when the number of covariates exceeds the number of observations (such as lasso), OLS fails in this case due to the $X^\top X$ matrix not being of full rank and hence not invertible. In cases with many covariates, but not more than the number of observations, estimation is unstable with many traditional statistical methods.

@@

Double robust methods for covariate adjustment are also motivated by the idea that many older methods ignore information present in the data. Methods tend to model either the outcome -- as in OLS regression and G-computation -- or only the treatment assignment -- as in propensity score matching or inverse probability weighting. Double robust methods, on the other hand, model both of these.

@@

Formally, the model can be written as the difference between an estimated outcome for treated units and the estimated outcome for untreated units (see the demonstration below):

>>>>

---
output:
  # bookdown::word_document2:
    # reference_docx: "word-template.docx"
  bookdown::pdf_document2:
    toc: no
    number_sections: yes
    pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
    #latex_engine: xelatex
always_allow_html: true
header-includes:
  #- \usepackage{setspace}\doublespace
  # - \usepackage[nolists, fighead, tabhead]{endfloat}
  # - \usepackage{endnotes}
  # - \let\footnote=\endnote
  # - \setlength{\headheight}{14.5pt}
  # - \setlength{\headheight}{13.6pt}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \lhead{N.I. Hoffmann}
  - \rhead{`r format(Sys.time(), '%B %e, %Y')`}
# - \newcommand{\beginsupplement}{\setcounter{table}{0}  
# \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} 
# \renewcommand{\thefigure}{A\arabic{figure}}}

@@

citeproc: no
#fontfamily: mathpazo
#fontsize: 11pt
# geometry: margin=.6in
indent: yes
link-citations: yes
linkcolor: blue
lang: 'en-US'

@@

title: "Double Robust, Flexible Adjustment Methods for Causal Inference: An Overview and an Evaluation"

author:  Nathan I. Hoffmann, Departments of Sociology and Statistics, UCLA
date: "`r format(Sys.time(), '%B %e, %Y')`"

abstract: "Double robust methods for flexible covariate adjustment in causal inference have proliferated in recent years. Despite their apparent advantages, these methods remain underutilized by social scientists. It is also unclear whether these methods actually outperform more traditional methods in finite samples. This paper has two aims: It is a guide to some of the latest methods in double robust, flexible covariate adjustment for causal inference, and it compares these methods to more traditional statistical methods and flexible \"single robust\" methods. It does this by using both simulated data where the treatment effect estimate is known, and then using comparisons of experimental and observational data from the National Supported Work Demonstration. Methods covered include Augmented Inverse Probability Weighting (AIPW), Targeted Maximum Likelihood Estimation (TMLE), and Double/Debiased Machine Learning (DML). Results suggest that some of these methods do outperform traditional methods in a wide range of simulations, but only slightly. In particular, the top performers are TMLE and AIPW in conjunction with flexible machine learning algorithms. But G-computation with the same flexible machine learning algorithms obtains almost identical results, and simple regression methods are nearly comparable in bias and are much more computationally efficient."

@@

library(SuperLearner)
library(broom)
library(knitr)
library(kableExtra)
library(here)
library(patchwork)
library(haven)
library(tidyverse)

@@

@@

  # pi_pred <-case_when(
  #   pi_pred < .01 ~ .01,
  #   pi_pred > .99 ~ .99,
  #   T ~ pi_pred)

@@

lalonde_setup <- function(dataset, sample, include_74 = F){

    list(y = dataset$re78,
                  z = ifelse(dataset$treat == 1, 'trt', 'ctl'),
                  x = dataset[,lalonde_variables],
                  y.1 = NA, y.0 = NA,
                  dataset = i,
                  set = sample,
                  size = nrow(dataset)) %>%
      return()
}

## Full lalonde
# sims <- list(
#   lalonde_setup(filter(lalonde_exp, treat == 0), sample = 'lalonde experimental original'),
#   lalonde_setup(lalonde_psid1_controls, 'lalonde PSID-1 original'),
#   lalonde_setup(lalonde_psid3_controls, 'lalonde PSID-3 original'),
#   lalonde_setup(lalonde_cps1_controls, 'lalonde CPS-1 original'),
#   lalonde_setup(lalonde_cps1_controls, 'lalonde CPS-3 original'),
#   lalonde_setup(filter(lalonde_exp_74, treat == 0), sample = 'lalonde experimental 74', include_74 = T),
#   lalonde_setup(lalonde_psid1_controls, 'lalonde PSID-1 74', include_74 = T),
#   lalonde_setup(lalonde_psid3_controls, 'lalonde PSID-3 74', include_74 = T),
#   lalonde_setup(lalonde_cps1_controls, 'lalonde CPS-1 74', include_74 = T),
#   lalonde_setup(lalonde_cps1_controls, 'lalonde CPS-3 74', include_74 = T))

## add lalonde ####
# read_csv(here('files_combined', 'lm_df.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde', 'lm_df.csv'))) %>%
#   write_csv(here('files', 'lm_df.csv'))
# 
# 
# read_csv(here('files_combined', 'psm_df.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde', 'psm_df.csv'))) %>%
#   write_csv(here('files', 'psm_df.csv'))
# 
# read_csv(here('files_combined', 'lin.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde', 'lin.csv'))) %>%
#   write_csv(here('files', 'lin.csv'))
# 
# read_csv(here('files_combined', 'gcomp.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde', 'gcomp.csv'))) %>%
#   write_csv(here('files', 'gcomp.csv'))
# 
# read_csv(here('files_combined/homemade', 'aipw.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde/homemade', 'aipw.csv'))) %>%
#   write_csv(here('files/homemade', 'aipw.csv'))
# 
# read_csv(here('files_combined/homemade', 'aipw_trunc.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde/homemade', 'aipw_trunc.csv'))) %>%
#   write_csv(here('files/homemade', 'aipw_trunc.csv'))
# 
# read_csv(here('files_combined/homemade', 'tmle.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde/homemade', 'tmle.csv'))) %>%
#   write_csv(here('files/homemade', 'tmle.csv'))
# 
# read_csv(here('files_combined/homemade', 'dml_ols_logit.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde/homemade', 'dml_ols_logit.csv'))) %>%
#   write_csv(here('files/homemade', 'dml_ols_logit.csv'))
# 
# read_csv(here('files_combined/homemade', 'dml_grf.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde/homemade', 'dml_grf.csv'))) %>%
#   write_csv(here('files/homemade', 'dml_grf.csv'))
# 
# read_csv(here('files_combined/homemade', 'dml_superlearner.csv')) %>%
#   bind_rows(read_csv(here('files_lalonde/homemade', 'dml_superlearner.csv'))) %>%
#   write_csv(here('files/homemade', 'dml_superlearner.csv'))
# 
# 
# readRDS(here('files_combined/homemade', 'ols_logit_pred.RDS')) %>%
#   append(readRDS(here('files_lalonde/homemade', 'ols_logit_pred.RDS'))) %>%
#   saveRDS(here('files/homemade', 'ols_logit_pred.RDS'))
# 
# readRDS(here('files_combined/homemade', 'grf_pred.RDS')) %>%
#   append(readRDS(here('files_lalonde/homemade', 'grf_pred.RDS'))) %>%
#   saveRDS(here('files/homemade', 'grf_pred.RDS'))
# 
# readRDS(here('files_combined/homemade', 'superlearner_pred.RDS')) %>%
#   append(readRDS(here('files_lalonde/homemade', 'superlearner_pred.RDS'))) %>%
#   saveRDS(here('files/homemade', 'superlearner_pred.RDS'))

@@

In causal inference, misspecification of underlying models can bias estimates of treatment effects [@hernan_2020_causal; @morgan_2015_counterfactuals]. There have been two important developments that attempt to overcome this. First, methodologists have developed machine learning methods that allow greater flexibility in estimation, adjusting for covariates in data-driven, complex ways [@balzer_2021_invited; @brand_2023_recent]. The second development is double robust methods [@bang_2005_doubly; @kang_2007_demystifying], which estimate two models: one for treatment exposure and another for the outcome. These models are robust to misspecification of either one of these models.

A number methods unifying these two developments have proliferated. These double robust methods for flexible covariate adjustment use machine learning methods to adapatively model the data generating processes. These models purportedly overcome the shortcomings of both traditional statistical methods and machine learning methods. Common statistical methods -- such as OLS regression and matching on propensity scores estimated from logistic regression -- have rigid functional form assumptions and fail to calculate stable estimates when the number of covariates is large relative to the number of observations. Machine learning methods, on the other hand, are often difficult to interpret. They can also suffer from overfitting, where the flexibility of the model becomes a weakness and predictions out-of-sample are poor, yet efforts to correct for overfitting can introduce regularization bias [@hastie_2009_elements]. Double robust methods with machine learning dispose of the constricting functional form assumptions of common statistical methods, and they correct for the regularization bias of flexible machine learning methods. They can also accommodate large numbers of covariates and produce easily interpretable treatment effect estimates. Despite their apparent advantages, these methods remain rarely utilized by social scientists. Part of the barrier has been lack of familiarity with these methods. It has also been unclear how these methods compare, or whether such methods actually perform better than traditional methods in finite samples.

This paper makes advances on these fronts. First, it is a guide to some of the latest methods in double robust, flexible covariate adjustment for causal inference, explaining the methods to a social scientist audience. Methods covered include Augmented Inverse Probability Weighting (AIPW), Targeted Maximum Likelihood Estimation (TMLE), and Double or Debiased Machine Learning (DML). This paper reviews the theory behind these methods as well as simple R implementations.

Second, this paper evaluates these methods. They are tested on simulations from @dorie_2019_automated, which cover a range of data-generating processes where ignorability holds -- i.e., there are no unmeasured confounders. They are then used to estimate experimental and observational effects from real-world data, the National Support for Work Demonstration (NSW) originally analyzed by @lalonde_1986_evaluating. In these evaluations, double robust methods are compared to "single robust" single robust methods. These include traditional or simpler statistical methods commonly used by social scientists: ordinary least squares (OLS) regression, matching on propensity scores estimated from logistic regression (PSM), and inverse probability weighting (IPW). They are also compared to two more flexible methods that may overcome the misspecification issues that double robust methods aim to overcome: G-computation and the Lin estimator [@lin_2013_agnostic].

Results from simulations where ignorability holds show that some double robust methods outperform traditional statistical methods, but only slightly. AIPW and TMLE perform the best, at least when used in conjunction with flexible machine learning algorithms, while DML does slightly worse than traditional methods. G-computation with flexible machine learning performs as well as the lowest-error double robust methods. Despite their lower average error, the computation times of these flexible methods are high. With its still relatively low error and much faster computation time, OLS remains a sensible choice as an estimator, and the Lin estimator -- which is as quick as standard OLS regression -- performs only slightly worse than the best double robust methods.

Results from the NSW study highlight the importance of ignorability. When observational samples are not comparable to the experimental sample, or when important covariates are not included, all methods -- double robust or traditional -- fail to recover experimental estimates. When appropriate covariates are included, most methods perform well. 
<!-- Exceptions include IPW, DML, and the Lin estimator (surprising given the simulation results), whose estimates tend to be less stable across samples.   -->

@@

Although previous work has compared different methods for covariate adjustment, there has not yet been a thorough evaluation and comparison of the recent and popular double robust methods of AIPW, TMLE, and DML for a social scientist audience.

Although some introductions to double robust methods exist, they do not discuss them in the context of covariate adjustment for causal inference, or their treatment is overly technical for a social scientist audience. For example, @kang_2007_demystifying provide an an excellent overview and evaluation of double robust methods, but in the context of missing data, and the authors consider AIPW but not TMLE or DML.  @bang_2005_doubly introduce double robust models for both causal inference and missing data, but their treatment is rather technical, and they only discuss AIPW.

Evaluations of double robust methods have focused on only one double robust method and compare to few traditional statistical methods. @dorie_2019_automated compare estimations from a number of different flexible methods, but these do not include AIPW or DML. @chatton_2020_gcomputation compare four methods -- G-computation, IPW, full matching, and TMLE -- but authors only consider one double robust method, and their focus is on omitted variable bias rather than determining which method is the most useful. @cousineau_2022_estimating evaluates the performance of optimization-based methods for causal inference, but these do not include the double robust methods covered in the current paper. @knaus_2022_double reviews double robust methods in an econometrics setting but does not compare them to traditional statistical methods for covariate adjustment.

An evaluation of multiple double robust methods that compares them to traditional statistical methods as well as flexible "single robust" methods is needed to understand just how practically useful these methods are for a social science audience. This paper endeavors to do this and to provide a gentle introduction to these methods.

@@

According to @bang_2005_doubly, double robust methods have their origins in missing data models. @robins_1994_estimation and @rotnitzky_1998_semiparametric developed augmented orthogonal inverse probability-weighted (AIPW) estimators in missing data models. Drawing on the fact that causal inference is fundamentally a missing data problem, @scharfstein_1999_adjusting showed that AIPW was double robust and extended to causal inference.

But @kang_2007_demystifying argue that double robust methods are older. They cite work by @cassel_1976_results, who proposed “generalized regression estimators” for population means from surveys where sampling weights must be estimated. Arguably, double robust methods go back even further than this. The form of double robust methods is similar to residual-on-residual regression, which dates back to the Frisch-Waugh-Lowell (FWL) theorem [@frisch_1933_partial; @lovell_1963_seasonal]:
$$\beta_D = \frac{\text{Cov}(\tilde Y_i, \tilde D_i)}{\text{Var}(\tilde D_i)}$$
where $\tilde D_i$ is the residual part of $D_i$ after regressing it on $X_i$, and $\tilde Y_i$ is the residual part of $Y_i$ after regressing it on $X_i$. This formulation writes the regression coefficient as composed of an outcome model ($\tilde Y_i$) and exposure model ($\tilde D_i$), the two models used in double robust estimators. Of the methods considered in this paper, double machine learning (DML) makes this connection most explicit by using residual-on-residual regression as part of its estimation strategy.

There are also links between double robust methods and matching with regression adjustment. This work goes back at least as far as @rubin_1973_use, who suggested that regression adjustment in matched data produces less biased estimates that either matching (exposure adjustment) or regression (outcome adjustment) do by themselves.

Today, double robust methods abound [e.g. @arkhangelsky_2021_doublerobust; @ratkovic_2023_relaxing; @sloczynski_2018_general; @dukes_2022_doubly; @kennedy_2023_semiparametric]. Although double robust methods exist for instrumental variables [@okui_2012_doubly; @wang_2018_bounded], difference-in-differences [@santanna_2020_doubly], longitudinal data [@yu_2006_double; @tran_2019_double], and other causal applications, this paper focuses on three of the most popular and foundational methods for covariate adjustment in a cross-sectional setting.

## Aims of Double Robust Methods
Double robust methods for covariate adjustment aim to overcome what many consider to be the shortcomings of both traditional statistical methods and flexible machine learning methods. Statistical methods that are popular with social scientists -- such as OLS regression and matching on propensity scores from logistic regression -- have two main weaknesses that double robust methods address. First, they assume simple (linear or transformed linear) functional forms. In the presence of highly nonlinear data generating processes, they may provide biased estimates. Second, these methods cannot handle large numbers of covariates relative to sample size. While some machine learning methods can produce estimates even when the number of covariates exceeds the number of observations (such as lasso), OLS fails in this case due to the $X^\top X$ matrix not being of full rank and hence not invertible. In cases with many covariates, but not more than the number of observations, estimation is unstable with many traditional statistical methods.

@@

Double robust methods attempt to overcome the downsides of both traditional and machine learning methods by incorporating flexible models into a framework that avoids overfitting and provides easily interpretable estimates. These methods are also motivated by the idea that many older methods ignore information present in the data. Methods tend to model either the outcome -- as in OLS regression and G-computation -- or only the treatment assignment -- as in propensity score matching or inverse probability weighting. Double robust methods, on the other hand, model both of these.

@@

Formally, the model can be written as the difference between an estimated outcome for treated units and an estimated outcome for untreated units (see the demonstration below):
